Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-20 introduction.tex
DSL syntax made same as OCaml syntax.

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support to allow
developers to verify the soundness of their applications, without
having to resort to low-level operational reasoning as they are forced
to do currently.  We develop a set of syntax-directed compositional
proof rules that enable the construction of correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules may discourage applications programmers from
using them, we also present an inference procedure based on these
rules that automatically verifies the weakest isolation level for a
transaction that still ensures its invariants are maintained.  The key
to inference is a novel formulation of database state (represented as
sets of tuples) as a monad, and in which database computations are
interpreted as state transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.  The paper makes the
following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
    embedded DSL for OCaml that treats SQL-based relational database
    operations (e.g., inserts, selects, deletes, joins, updates, etc.)
    as monadic computations over an abstract database state.
 \item We develop an operational semantics and a compositional
   rely-guarantee style proof system for this language capable of
   relating high-level application invariants to database state,
   parameterized by a weak isolation semantics that selectively
   exposes the visibility of these operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

>>>>

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support to allow
developers to verify the soundness of their applications, without
having to resort to low-level operational reasoning as they are forced
to do currently.  We develop a set of syntax-directed compositional
proof rules that enable the construction of correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules may discourage applications programmers from
using them, we also present an inference procedure based on these
rules that automatically verifies the weakest isolation level for a
transaction that still ensures its invariants are maintained.  The key
to inference is a novel formulation of database state (represented as
sets of tuples) as a monad, and in which database computations are
interpreted as state transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.  The paper makes the
following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
  \item We analyze properties of weak isolation in the context of a
    DSL embedded in OCaml that treats SQL-based relational database
    operations (e.g., inserts, selects, deletes, updates, etc.) as
    computations over an abstract database state.
  \item We develop an operational semantics and a compositional
    rely-guarantee style proof system for this language capable of
    relating high-level application invariants to database state,
    parameterized by a weak isolation semantics that selectively
    exposes the visibility of these operations to other transactions.
  \item We devise an inference algorithm capable of discovering the
    weakest isolation level that is sound with respect to a
    transaction's high-level consistency requirements. The algorithm
    interprets database operations as state transformers expressed in
    a language amenable for translation into a decidable fragment of
    first-order logic, and is thus suitable for automated verification
    using off-the-shelf SMT solvers.
  \item We present details of an implementation along with an
    evaluation study on real database benchmarks that justify our
    approach, and demonstrate the utility of our inference mechanism.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

### 2017-10-20 motivation.tex
DSL syntax made same as OCaml syntax.

We present our ideas in the context of an embedded DSL in OCaml that
we have developed, which offers a \C{DB} monad to define and compose
database computations given in terms of SQL operations.  Besides the
usual \C{bind} and \C{return} operators, the monad includes an
\C{atomically} combinator that executes a database computation as an
atomic transaction and returns the result.

\begin{figure}
\centering
\begin{ocaml}
let new_order (d_id, c_id, item_reqs) = atomically do
  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
  let o_id = dist.d_next_o_id in
      SQL.update(* UPDATE *) District 
                (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
                (* WHERE *)(fun d -> d.d_id = d_id );
      SQL.insert(* INSERT INTO *) Order (* VALUES *){o_id=o_id;  
                o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
      foreach item_reqs @@ fun item_req -> do
        stk <- SQL.select1(* SELECT * FROM *) Stock 
                  (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
                                      && s.s_d_id = d_id)(* LIMIT 1 *); 
        let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
                    then stk.s_qty - item_req.ol_qty 
                    else stk.s_qty - item_req.ol_qty + 91 in
            SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
                             (fun s -> s.s_i_id = item_req.ol_i_id);
            SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
                                   ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An executing RC transaction is isolated from
\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
allowed to witness the writes of concurrent transactions as soon as
they are committed. Thus, with two concurrent instances of the
\C{new\_order} transaction (call them $T_1$ and $T_2$), both
concurrently placing new orders for different customers under the same
district (\C{d\_id}), RC isolation allows the execution shown in
Fig.~\ref{fig:new_order_exec}.

@@

Clearly, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes} - both concurrent transactions' writes are present
in the final state of the database.  Program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

@@

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state monadically,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
records, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

% \begin{figure}[!t]
% \centering
% %
% \begin{subfigure}[b]{0.46\textwidth}
% \begin{ocaml}
% Set s' = $\emptyset$;
% foreach x in s {
%   s'.add(f(x)); 
% }
% \end{ocaml}
% \caption{}
% \end{subfigure}
% %
% \begin{subfigure}[b]{0.5\textwidth}
% \begin{ocaml}
% let s' = ref [];
% foreach s (fun x -> s' := x::(!s'));
% \end{ocaml}
% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
% \end{subfigure}
% \caption{Caption place holder}
% %
% \caption{New versions are created from existing versions either
% through \C{push} or \C{merge}.}
% \label{fig:syntactic-ancestors}
% \end{figure}
% let s' = ref (Set.empty) in
% foreach xs @@ fun x -> 
%   begin
%     s' := Set.union !s' @@ 
%             Set.map_selected s (fun y -> y<x)
%                                (fun y -> y+x);
%     s' := Set.add !s' x;
%   end
\begin{figure}[!h]
\begin{ocaml}
foreach item_reqs @@ fun item_req -> do
  SQL.update Stock (fun s -> {s with s_qty = k1}) 
                   (fun s -> s.s_i_id = item_req.ol_i_id);
  SQL.insert Order_line {ol_o_id=k2; ol_d_id=k3; 
                         ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
\end{ocaml}
\caption{Foreach loop from Fig.~\ref{fig:new_order_code}}
\label{fig:foreach_code}
\end{figure}

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
        s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
           {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                       \C{s\_d\_id}=s.\C{s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to the local database has not previously been
  updated.}

@@

\end{array}
\end{smathpar}
Observe that both transformers are of the form $\T(\delta,\Delta) =
\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
set of records added  to the transaction-local database ($\stl$). Let
$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
shown above. The state transformation induced by the loop body in
Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
As we demonstrate, the advantage of inferring such transformers is
that we can now make use of a semantics-preserving translation from
the domain of sets equipped with $\bind$ to a decidable fragment of
first-order logic, allowing us to leverage SMT solvers for automated
proofs without having to infer potentially complex thread-local
invariants or intermediate assertions.  Sec.~\ref{sec:inference}
describes this translation. In the exposition thus far, we assumed
$\Delta$ remains invariant, which is clearly not the case when we
admit concurrency.  Necessary concurrency extensions of the state
transformer semantics to deal with interference is also covered in
Sec.~\ref{sec:inference}.  Before presenting the transformer
semantics, we first focus our attention in the following two sections
on the theoretical foundations for weak isolation, upon which this
semantics is based.

>>>>

We present our ideas in the context of a DSL embedded in OCaml that
manages an abstract database state, and allows it be manipulated via a
well-defined \C{SQL} interface. Arbitrary database computations can be
built around this interface, which can then be run as transactions
using the \C{atomically\_do} combinator provided by the DSL.

\begin{figure}
\centering
\begin{ocaml}
let new_order (d_id, c_id, item_reqs) = atomically_do @@ fun () ->
  let dist = SQL.select1 District (fun d -> d.d_id = d_id) in
  let o_id = dist.d_next_o_id in
  SQL.update(* UPDATE *) District 
            (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
            (* WHERE *)(fun d -> d.d_id = d_id );
  SQL.insert(* INSERT INTO *) Order (* VALUES *){o_id=o_id;  
            o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
  foreach item_reqs @@ fun item_req ->
    let stk = SQL.select1(* SELECT * FROM *) Stock 
              (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
                                  && s.s_d_id = d_id)(* LIMIT 1 *); 
    let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
                then stk.s_qty - item_req.ol_qty 
                else stk.s_qty - item_req.ol_qty + 91 in
        SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
                         (fun s -> s.s_i_id = item_req.ol_i_id);
        SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
                               ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction when executed under
\emph{Read Committed} (RC) isolation level, the default isolation
level in 8 of the 18 databases studied in~\cite{bailishotos}.
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
An executing RC transaction is isolated from \emph{dirty writes},
i.e., writes of uncommitted transactions, but is allowed to witness
the writes of concurrent transactions as soon as they are committed.
Thus, with two concurrent instances of the \C{new\_order} transaction
(call them $T_1$ and $T_2$), both concurrently placing new orders for
different customers under the same district (\C{d\_id}), RC isolation
allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

Clearly, this example does not exhibit any of the anomalies that
\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
al~\cite{berenson} characterize isolation levels in terms of the
anomalies they \emph{admit}. For e.g., RC is characterized by
\emph{lost writes} because it admits the anomaly.}. For instance,
there are no \emph{lost writes} - both concurrent transactions' writes
are present in the final state of the database.  Program analyses that
aim to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

@@

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state abstractly,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
records, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

% \begin{figure}[!t]
% \centering
% %
% \begin{subfigure}[b]{0.46\textwidth}
% \begin{ocaml}
% Set s' = $\emptyset$;
% foreach x in s {
%   s'.add(f(x)); 
% }
% \end{ocaml}
% \caption{}
% \end{subfigure}
% %
% \begin{subfigure}[b]{0.5\textwidth}
% \begin{ocaml}
% let s' = ref [];
% foreach s (fun x -> s' := x::(!s'));
% \end{ocaml}
% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
% \end{subfigure}
% \caption{Caption place holder}
% %
% \caption{New versions are created from existing versions either
% through \C{push} or \C{merge}.}
% \label{fig:syntactic-ancestors}
% \end{figure}
% let s' = ref (Set.empty) in
% foreach xs @@ fun x -> 
%   begin
%     s' := Set.union !s' @@ 
%             Set.map_selected s (fun y -> y<x)
%                                (fun y -> y+x);
%     s' := Set.add !s' x;
%   end
\begin{figure}[!h]
\begin{ocaml}
foreach item_reqs @@ fun item_req ->
  SQL.update Stock (fun s -> {s with s_qty = k1}) 
                   (fun s -> s.s_i_id = item_req.ol_i_id);
  SQL.insert Order_line {ol_o_id=k2; ol_d_id=k3; 
                         ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
\end{ocaml}
\caption{Foreach loop from Fig.~\ref{fig:new_order_code}}
\label{fig:foreach_code}
\end{figure}

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

@@

\end{array}
\end{smathpar}
Observe that both transformers are of the form $\T(\delta,\Delta) =
\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
set of records added  to the transaction-local database ($\stl$). Let
$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
shown above. The state transformation induced by the loop body in
Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{foreach} combinator becoms set monad's bind
($\bind$) combinator.  As we demonstrate, the advantage of inferring
such transformers is that we can now make use of a
semantics-preserving translation from the domain of sets equipped with
$\bind$ to a decidable fragment of first-order logic, allowing us to
leverage SMT solvers for automated proofs without having to infer
potentially complex thread-local invariants or intermediate
assertions.  Sec.~\ref{sec:inference} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invariant, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics to deal with
interference is also covered in Sec.~\ref{sec:inference}.  Before
presenting the transformer semantics, we first focus our attention in
the following two sections on the theoretical foundations for weak
isolation, upon which this semantics is based.
